Fixing Vulnerabilities Potentially Hinders Maintainability

### 2020-09-03 paper/emse20.tex
hot fixes

\newcommand\Sof[1]{\nb{Sofia}{red}{#1}}
\newcommand\Luis[1]{\nb{Luis}{mygreen}{#1}}
\newcommand\Rui[1]{\nb{Rui}{blue}{#1}}

@@

ISO describes \textit{software maintainability} as ``the degree of 
effectiveness and efficiency with which a software product or system 
can be modified to improve it, correct it or adapt it to changes in 
environment, and in requirements'' on software quality ISO/IEC 
25010~\cite{iso:2011}. Thereby, maintainable security may be 
defined, briefly, as the degree of effectiveness and efficiency with 
which software can be changed to mitigate a security 
vulnerability---corrective maintenance.
% (hopefully, without introducing new ones). 
However, many developers still lack knowledge on the best 
practices to deliver and maintain secure and high-quality 
software~\cite{Pothamsetty:2005:SEL:1107622.1107635,8077802}. In a world where zero-day vulnerabilities are constantly 
emerging, mitigation needs to be fast and efficient. Therefore, it 
is important to write maintainable code to support the production of 
more secure software---maintainable code is faster 
to mitigate---and, prevent the introduction of new vulnerabilities 
in the codebases.
% Unmaintainable code is hard to test, analyze and re-use. Therefore,
% it is important to write maintainable code to support the production
% of more secure software and prevent
% the introduction of new vulnerabilities in the code bases.
%
Novel tools have been built to automatically detect software 
vulnerabilities (e.g., FindBugs, Infer and more). Despite these 
efforts, improving software security is not a trivial task and 
requires implementing patches that might affect software 
maintainability. Our suspicion is that some of these patches may 
have a negative impact on the software maintainability and, 
possibly, even be the cause of the introduction of new 
vulnerabilities---harming software reliability and introducing 
technical debt. Therefore, in this paper, we present an empirical 
study on the impact of patches of vulnerabilities on software 
maintenance across open-source software.

@@

This research performs the following main contributions:
%
\begin{itemize}
  \item Evidence that supports the trade-off between security and 
  maintainability: developers may be hindering software 
  maintainability while patching vulnerabilities.
	\item An empirical study on the impact of security patches on 
	software maintainability (per guideline, severity, weakness and 
	programming language).
	\item A replication package with all scripts and data created to 
	perform the empirical evaluation, for reproducibility. Available 
	online: \url{https://figshare.com/s/4861207064900dfb3372}.
\end{itemize}
%
This paper is structured as follows: Section~\ref{sec:motivation} 
introduces an example of a security patch of a known vulnerability 
found in the protocol implementation of 
OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
contains open-source implementations of the SSL and TLS cryptographic
protocols. Repository available at 
\url{https://github.com/openssl/openssl} (Accessed on \today{})}; 
Section~\ref{sec:methodology} describes the methodology used to 
answer the research questions; Section~\ref{sec:results} presents 
the results and discusses their implications; 
Section~\ref{sec:implications} elaborates on the implications
developers should consider in the future; Section~\ref{sec:threats} 
enumerates the threats to the validity of this study; 
Section~\ref{sec:rw} describes the different work and existing
literature in the field of study; and, finally, 
Section~\ref{sec:conclusions} concludes the main findings and 
elaborates on future work.

@@

In this section, we discuss the methodology used to measure
the impact of security patches on the maintainbility of open-source
software. The methodology comprises the following steps, as 
illustrated in Fig.~\ref{fig:met}.
%
\begin{figure}[h]
	\centering 	
    \begin{adjustwidth}{0.5cm}{-1cm}  
	\includegraphics[width=0.9\textwidth]{figures/methodology.pdf}

 	\caption{Study Methodology}
	\label{fig:met}
	 \end{adjustwidth}
\end{figure}
%
\begin{enumerate}
	\item Combine the datasets from related work that classifies
the activities of developers addressing security-oriented patches~\cite{Reis:2017:IJSSE,10.1109/MSR.2019.00064}.
%
	\item
	Extract relevant data (e.g., owner and name of
	the repository, sha key of the vulnerable version, sha key of the fixed version) to the experiments and 
	code bases from the combined dataset containing $1335$ security
	patches collected from open-source software available on
	GitHub.
%
	\item A baseline of regular commits was randomly collected from the list of
	projects of the final dataset to evaluate the impact of regular commits on the
	maintainability of open-source software.
%
  \item Use the Software Improvement Group's\footnote{SIG's website: \url{https://www.sig.eu/} 
  (Accessed on \today{})} web-based source code analysis
  service \emph{Better Code Hub} (BCH)\footnote{BCH's website:
  \url{https://bettercodehub.com/} (Accessed on \today{})} to quantify maintainability
  for both security and regular commits. The code bases that resemble to the commits
  of our datasets are collected before the analysis performed by BCH.
\end{enumerate}
%

%
\subsection{Datasets}
%

We use a combined dataset of security vulnerabilities which is 
the outcome of mining and manually inspect a total of $265$ GitHub projects for both
datasets. Reis and Abreu 
($2017$) mined open-source
software aiming at the extraction of real examples---created by real
developers---to test and assess the performance of static analysis tools~\cite{Reis:2017:IJSSE} since
using hand-seeded test cases or mutations could lead to misleading assessments
of the capabilities of the tools~\cite{just2014mutants}. The study yielded a
dataset of $711$ test cases for $16$ different security vulnerability types. Some of the vulnerability types
are based on the OWASP Top $10$ of $2013$~\cite{oswap:2013} and OWASP Top $10$ of
$2017$~\cite{oswap:2017}. Each test case of the
dataset is a triplet: the commit before the patching, the commit responsible
for the patching, and the snippets of code that differ from one version to
another (typically, called \textit{diffs})---where one can easily review the
code used to fix the security flaw. 
%

Ponta et al. ($2019$) tracked the \url{pivotal.io} website for vulnerabilities 
from $2014$ to $2019$. For each new vulnerability, the authors manually searched 
for references to commits involved in the patch on the National Vulnerability 
Database (NVD) website. However, $70\%$
of the vulnerabilities did not have any references to commits. Thus, the authors
used their expertise to locate the commits in the repositories. This technique 
yielded a database of $624$ patches~\cite{10.1109/MSR.2019.00064}. The original 
dataset contains $1282$ commits. One patch can have multiple commits assigned.
To fit the dataset in our methodology, we search for the first and last commits
used to patch the vulnerability. We assumed the last commit of the patch as 
the fix (\emph{sha}) and the parent of the first commit as the vulnerable version 
(\emph{sha-p}).
%

@@

%
% \begin{itemize}
%     \item \textbf{CWE-89: Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases).} When developers do not keep untrusted data
%     separate from SQL queries. If an attacker sends a command that
%     exploits the syntax of the SQL interpreter, then SQL injection attack is possible.
% %
%     \item \textbf{CWE-611: Improper Restriction of XML External Entity Reference (24 cases).}
% 	XML documents are exchanged through the web containing entities with URIs
% 	that resolve to local/external files. Thus, when XML parsers are not well configured,
% 	attackers have allowed to directly those files.
% %
%     \item \textbf{CWE-20: Improper Input Validation (76 cases).} When developers
% 	forget to validate the inputs of an application, attackers may have control
% 	of the control/data flow of the program.
% %
%     \item \textbf{CWE-200: Information Exposure (39 cases).} Application's logs are one way of intentional/unintentional disclosure information
% 	to attackers. Many times attackers get access to logs when sould not be authorized to.
% %
% 	\item \textbf{CWE-352: Cross-Site Request Forgery/CSRF (22 cases).} Poor session tokens generation
%     and management usually allow attackers to send forged HTTP requests
%     including authentication information from the victim to the vulnerable web
%     application.
% %
% 	\item \textbf{CWE-264: Permissions, Privileges, and Access Controls (59 cases).} Incorrectly
% 	    implemented functionalities related to authentication and session
% 	    management, allowing an attacker to gain access to session tokens,
% 	    passwords, keys, and other sensitive data.
% %
% 	\item \textbf{CWE-399: Resource Management Errors (22 cases).} Resource management issues are found more frequently
% 		in programming languages that do not manage memory automatically (e.g., C/C++
% 	    and Objective-C), i.e., where developers are responsible for
% 	    handling it instead. It is one of the main causes of DoS attacks.
% %
% 	\item \textbf{CWE-79: Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)} Lack of proper validation or escaping
% 	    allows attackers to submit untrusted data to web browsers through malicious
% 	    scripts that can hijack the user sessions or redirect the user to malicious
% 	    sites.
% %
% 	\item \textbf{CWE-22: Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases).}  When developers forget to neutralize special elements within the pathnames to files/directories, attackers
% 	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
% 	or entire repositories.
% %
% 	\item \textbf{Miscellaneous (673 cases)} This pattern comprises
% 		several other security patches that do not have a CWE assigned and patterns that
% 		do not satisfy the Wilcoxon test size requirement of more than $20$ patches.
% \end{itemize}
%
\subsection{Security vs. Baseline Commits}
%
Previous studies attempted to measure the impact of regular refactorings on
open-source software maintainability~\cite{HEGEDUS2018313}. However, there is no
previous work focused on comparing the impact of security patches with
regular refactorings on maintainability.
We analyze the maintainability of regular commits (i.e., commits not related
to security fixes) and use them as a baseline.
The baseline dataset is generated from the security commits dataset.
For each
security commit, a commit with an approximate size is fetched. Due to the complexity 
of some patches, it was not possible to find patches with the exact number of added and 
deleted lines. Our search uses the patch \emph{diff} to look for a new similar \emph{diff}. 
The algorithm integrates the following steps:

@@

	\item How this is a problem or has an impact to the software project? 
  What is the message to developers? (clarify)

@@

  \item Review the secure programming languages implication. Check 

@@

  \item [!] \textit{It did take me several readings to understand the paper, especially the approach} - 
  what to do with this @rui?
      \item Ehmm... strange... we need to improve the approach then. Thoughts?

@@

  \item \textit{authors write "...important to write maintainable code to support the production
  of more secure software...". Relationship is not clear nor is it backed by some evidence} -> 
  Improve this.
  \item Drop the word patch. Security patches are not patches. Refactorings is perfective
  or preventive maintenance, whereas patching a vulnerability is corrective maintenance.

@@

  \item \textit{Authors state that they assume for the Pontas dataset that the last commit related 
  to patching being the actual patch while the parent of the first patch-related commit is assumed 
  to be the vulnerable version. However, nothing is said whether this assumption holds all the time. 
  So, how did authors validate their assumption? This is missing, but should be added.} - Improve this.

>>>>

\newcommand\Sofia[1]{\nb{Sofia}{red}{#1}}
\newcommand\Luis[1]{\nb{Luis}{mygreen}{#1}}
\newcommand\Rui[1]{\nb{Rui}{blue}{#1}}

@@

ISO describes \textit{software maintainability} as ``the degree of 
effectiveness and efficiency with which a software product or system 
can be modified to improve it, correct it or adapt it to changes in 
environment, and in requirements'' on software quality ISO/IEC 
25010~\cite{iso:2011}. Thereby, maintainable security may be 
defined, briefly, as the degree of effectiveness and efficiency with 
which software can be changed to mitigate a security 
vulnerability---corrective maintenance.
% (hopefully, without introducing new ones). 
However, many developers still lack knowledge on the best 
practices to deliver and maintain secure and high-quality 
software~\cite{Pothamsetty:2005:SEL:1107622.1107635,8077802}. In a 
world where zero-day vulnerabilities are constantly emerging, 
mitigation needs to be fast and efficient. Therefore, it 
is important to write maintainable code to support the production of 
more secure software---maintainable code is less complex and, 
consequently, less prone to
vulnerabilities~\cite{shin2010evaluating,10.1145/1774088.1774504})---
and, prevent the introduction of new vulnerabilities.
% Unmaintainable code is hard to test, analyze and re-use. Therefore,
% it is important to write maintainable code to support the production
% of more secure software and prevent
% the introduction of new vulnerabilities in the code bases.
%
Novel tools have been built to automatically detect software 
vulnerabilities (e.g., FindBugs, Infer and more). Despite these 
efforts, improving software security is not a trivial task and 
requires implementing patches that might affect software 
maintainability. Our suspicion is that some of these patches may 
have a negative impact on the software maintainability and, 
possibly, even be the cause of the introduction of new 
vulnerabilities---harming software reliability and introducing 
technical debt. Therefore, in this paper, we present an empirical 
study on the impact of patches of vulnerabilities on software 
maintenance across open-source software.

@@

This research performs the following main contributions:
%
\begin{itemize}
	\vspace{-0.3em}
  \item Evidence that supports the trade-off between security and 
  maintainability: developers may be hindering software 
  maintainability while patching vulnerabilities.
	\item An empirical study on the impact of security patches on 
	software maintainability (per guideline, severity, weakness and 
	programming language).
	\item A replication package with all scripts and data created to 
	perform the empirical evaluation, for reproducibility. Available 
	online: \url{https://figshare.com/s/4861207064900dfb3372}.
\end{itemize}
%
This paper is structured as follows: Section~\ref{sec:motivation} 
introduces an example of a security patch of a known vulnerability 
found in the protocol implementation of 
OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
contains open-source implementations of the SSL and TLS cryptographic
protocols. Repository available at 
\url{https://github.com/openssl/openssl} (Accessed on \today{})}; 
Section~\ref{sec:methodology} describes the methodology used to 
answer the research questions; Section~\ref{sec:results} presents 
the results and discusses their implications; 
Section~\ref{sec:implications} elaborates on the implications
developers should consider in the future; Section~\ref{sec:threats} 
enumerates the threats to the validity of this study; 
Section~\ref{sec:rw} describes the different work and existing
literature in the field of study; and, finally, 
Section~\ref{sec:conclusions} concludes the main findings and 
elaborates on future work.

@@

In this section, we discuss the methodology used to measure
the impact of security patches on the maintainbility of open-source
software. The methodology comprises the following steps, as 
illustrated in Fig.~\ref{fig:met}.
%
\begin{enumerate}
	\item Combine the datasets from related work that classifies
	the activities of developers addressing security-oriented 
	patches~\cite{Reis:2017:IJSSE,10.1109/MSR.2019.00064}.
%
	\item
	Extract relevant data (e.g., owner and name of
	the repository, sha key of the vulnerable version, sha key of 	
	the fixed version and codebases) from the combined dataset 	
	containing $1335$ security patches collected from open-source 	
	software available on GitHub.
%
	\item A baseline of regular commits was randomly collected from 
	the list of projects of the final dataset to evaluate the impact 
	of regular commits on the maintainability of open-source 
	software.
%
  \item Use the Software Improvement Group's\footnote{SIG's website: 
  \url{https://www.sig.eu/} (Accessed on \today{})} web-based source 
  code analysis service \emph{Better Code Hub} (BCH)\footnote{BCH's 
  website: \url{https://bettercodehub.com/} (Accessed on \today{})} 
  to quantify maintainability for both security and regular commits. 
  The codebases that resemble to the commits of our datasets are 
  collected before the analysis performed by BCH.
\end{enumerate}
%
\begin{figure}[h]
	\centering 	
    \begin{adjustwidth}{0.5cm}{-1cm}  
	\includegraphics[width=0.9\textwidth]{figures/methodology.pdf}
 	\caption{Study Methodology}
	\label{fig:met}
	 \end{adjustwidth}
\end{figure}
%
\subsection{Datasets}
%

We use a combined dataset of security vulnerabilities which is 
the outcome of mining and manually inspecting a total of $265$ 
GitHub projects for both datasets. Reis and Abreu 
($2017$) mined open-source software aiming at the extraction of 
real---created by developers---samples of security vulnerabilities 
to test and assess the performance of static analysis 
tools~\cite{Reis:2017:IJSSE} since using hand-seeded test cases or 
mutations can lead to misleading assessments of the capabilities of 
the tools~\cite{just2014mutants}. The study yielded a dataset of 
$711$ test cases for $16$ different security vulnerability types. 
Some of the vulnerability types are based on the OWASP Top $10$ of 
$2013$~\cite{oswap:2013} and OWASP Top $10$ of 
$2017$~\cite{oswap:2017}. Each test case of the dataset is a 
triplet: the commit before the patching, the commit responsible
for the patching, and the snippets of code that differ from one 
version to another (typically, called \textit{diffs})---where one 
can easily review the code used to fix the security flaw. 
%

Ponta et al. ($2019$) used a different approach and tracked the 
\url{pivotal.io} website for vulnerabilities from $2014$ to $2019$. 
For each new vulnerability, the authors manually searched for 
references to commits involved in the patch on the National 
Vulnerability Database (NVD) website. However, $70\%$ of the 
vulnerabilities did not have any references to commits. Thus, the 
authors used their expertise to locate the commits in the 
repositories. This technique yielded a dataset of $624$ 
patches~\cite{10.1109/MSR.2019.00064}. The original dataset contains 
$1282$ commits. One patch can have multiple commits assigned.
To fit the dataset in our methodology, we located the first and 
last commits used to patch the vulnerability. For these cases, we 
used the GitHub API to retrieve automatically the dates of the 
commits. Then, for each patch, the group of commits was ordered from 
the oldest commit to the newest one. We assumed the last commit 
(newest one) as the fix (\emph{sha}) and the parent of the 
first commit (oldest commit) as the vulnerable version 
(\emph{sha-p}). 
%

@@

%
% \begin{itemize}
%     \item \textbf{CWE-89: Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases).} When developers do not keep untrusted data
%     separate from SQL queries. If an attacker sends a command that
%     exploits the syntax of the SQL interpreter, then SQL injection attack is possible.
% %
%     \item \textbf{CWE-611: Improper Restriction of XML External Entity Reference (24 cases).}
% 	XML documents are exchanged through the web containing entities with URIs
% 	that resolve to local/external files. Thus, when XML parsers are not well configured,
% 	attackers have allowed to directly those files.
% %
%     \item \textbf{CWE-20: Improper Input Validation (76 cases).} When developers
% 	forget to validate the inputs of an application, attackers may have control
% 	of the control/data flow of the program.
% %
%     \item \textbf{CWE-200: Information Exposure (39 cases).} Application's logs are one way of intentional/unintentional disclosure information
% 	to attackers. Many times attackers get access to logs when sould not be authorized to.
% %
% 	\item \textbf{CWE-352: Cross-Site Request Forgery/CSRF (22 cases).} Poor session tokens generation
%     and management usually allow attackers to send forged HTTP requests
%     including authentication information from the victim to the vulnerable web
%     application.
% %
% 	\item \textbf{CWE-264: Permissions, Privileges, and Access Controls (59 cases).} Incorrectly
% 	    implemented functionalities related to authentication and session
% 	    management, allowing an attacker to gain access to session tokens,
% 	    passwords, keys, and other sensitive data.
% %
% 	\item \textbf{CWE-399: Resource Management Errors (22 cases).} Resource management issues are found more frequently
% 		in programming languages that do not manage memory automatically (e.g., C/C++
% 	    and Objective-C), i.e., where developers are responsible for
% 	    handling it instead. It is one of the main causes of DoS attacks.
% %
% 	\item \textbf{CWE-79: Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)} Lack of proper validation or escaping
% 	    allows attackers to submit untrusted data to web browsers through malicious
% 	    scripts that can hijack the user sessions or redirect the user to malicious
% 	    sites.
% %
% 	\item \textbf{CWE-22: Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases).}  When developers forget to neutralize special elements within the pathnames to files/directories, attackers
% 	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
% 	or entire repositories.
% %
% 	\item \textbf{Miscellaneous (673 cases)} This pattern comprises
% 		several other security patches that do not have a CWE assigned and patterns that
% 		do not satisfy the Wilcoxon test size requirement of more than $20$ patches.
% \end{itemize}
%
\subsection{Security vs. Baseline Commits}
%
Previous studies attempted to measure the impact of regular commits 
on open-source software maintainability~\cite{HEGEDUS2018313}. 
However, there is no previous work focused on comparing the impact 
of security patches with regular commits on maintainability.
We analyze the maintainability of regular commits (i.e., commits not 
related to security patches) and use them as a baseline.
The baseline dataset is generated from the security commits dataset.
For both vulnerable/sha-p and patch/sha commits, a commit 
with an approximate size is fetched. Due to the complexity of some 
patches, it was not possible to find patches with the exact number 
of added and deleted lines. Our search uses the patch \emph{diff} to 
look for a \emph{diff} of the same size. The algorithm integrates 
the following steps:

@@

	% \item How this is a problem or has an impact to the software project?
	%   What is the message to developers? (clarify)

@@

  \item Implications: Review the secure programming languages implication. Check 

@@

  % \item [!] \textit{It did take me several readings to understand the paper, especially the approach} -
 %  what to do with this @rui?
 %    \begin{itemize}
 %      \item Ehmm... strange... we need to improve the approach then. Thoughts?
 %    \end{itemize}

@@

  % \item \textit{authors write "...important to write maintainable code to support the production
  % of more secure software...". Relationship is not clear nor is it backed by some evidence} ->
  % Improve this.
  % \item Drop the word patch. Security patches are not patches. Refactorings is perfective
  % or preventive maintenance, whereas patching a vulnerability is corrective maintenance.

@@

  % \item \textit{Authors state that they assume for the Pontas dataset that the last commit related
  % to patching being the actual patch while the parent of the first patch-related commit is assumed
  % to be the vulnerable version. However, nothing is said whether this assumption holds all the time.
  % So, how did authors validate their assumption? This is missing, but should be added.} - Improve this.
